Oak resilience to drought and land-use show how the ecological and geographical rear-edges do not necessarily meet today

### 2019-08-26 man/ecosystems/ms_ecosystems.Rmd
update contents

# Discussion
## Land-use legacies shape forests and sensitivity to climate change at the rear-edge

The Mediterranean landscapes, even at high elevations, have suffered a strong transformation driven by human activities throughout history [@GarciaAlix2017]. Forest clearings, firewood extraction, charcoal exploitations and mining (these were strongly suggested in the disturbance chronology in SJ), have strongly impacted on the forest resources at Sierra Nevada as can be derived from the results of the review of historical documents (Table S4). Indeed a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation has been documented in this mountain region [@JimenezOlivencia2015]. 
Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland) [@JimenezOlivencia2015]. It has been described a sequentially distribution of the land uses along the elevational gradient in the northern slopes of Sierra Nevada (*e.g.* SJ site): grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops [@JimenezOlivencia2015]. Other activities (like mining) should have determined the forest structure at that northern location, since SJ is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history. On the other hand, woodland areas of the southern slopes (*e.g.* CA site) are mixed with a greater percentage of croplands along the elevational gradient where oaks occur [@JimenezOlivencia2015]. Firewood, charcoal extraction and acorns have been resources intensively exploited in southern sites, until at least the middle of the 20th century when these activities sharply decreased mainly due to rural abandonment and use of gas and fossil fuels [@Valbuena2013; @JimenezOlivencia2015].

Some of the previous historical records of land-use were expressed in our growth chronologies. The release event at 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral [@Titos1990]. This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth the remnant trees included in the northern site (Figure S4). The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 5, S4). There are no records of forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest] and no record for any logging  during 1995 - 2000 period (F.J. Cano-Manuel *personal communications*). We speculate that this might be related with a natural drought-induced mortality event after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004]. On the other hand, the strong positive correlations of SPEI with tree-growth for this site show a high sensitivity to water availability [@GeaIzquierdo2014]. For southern sites, the only positive release event showed at the earliest years for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] documented in this site [@Valbuena2013].

As we inferred from our results, the management history can conditioned tree-growth and resilience of tree species, highlighting the importance of land-use legacies in Mediterranean forests [@Navarro2013; @DoblasMiranda2017]. Nevertheless, we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource. This pattern concurs with a forest expansion into marginal abandoned croplands and an increase in the tree-cover [@CamachoOlmedo2002].

@@

Severe drought negatively affects both primary and secondary growth of *Q. pyrenaica* forests. This was expressed by the observed reduction in greenness and tree-growth in response to the 2005 and 2012 drought events, and the consistent radial growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014]. Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018]. Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g.* 2005 and 2012 were considered two of the worst recorded drought; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* for the last 16 years, which is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather humid decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto]. For tree-growth, we also observed positive trends in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along its distribution range only at high elevation sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold limited high elevation sites [@Salzer2009; @GeaIzquierdo2014]. However, the observed pattern for our sites (a rear-edge) differs from the declining trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

Despite the 2012 drought event was more severe and intense than 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012. This could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] occurring earlier than the 2005 drought. The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q. pyrenaica* forests (Figure S7). Thus, this highlights the importance of the timing of the drought as a key factor determining tree recovery after drought [@Huang2018], as also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b]. For tree-growth, the highest values of resilience were obtained for the most severe events, namely 1995 and 1999 (Table S3). In addition, we observed a positive relation between the tree-growth resilience and drought severity (Figures 6). This result seems to contradict our initial hypothesis in which we predicted low resilience values for this oak species in their rear-edge after extreme drought events. Yet, the high resilience values obtained for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality (F.J. Navarro *personal communications*), which would show the influence of stand dynamics partly shadowing interpretation of individual resilience estimated from individual growth series.

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. A high vulnerability to drought is often assumed for populations located at their rear-edge [@MartinezVilalta2018]. However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q. pyrenaica* and high values of resilience to severe droughts. Our findings are in agreement with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017]. @MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations. The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013], seem to be indicative of high local adaptation of this oak. In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak. A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018]. In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q. pyrenaica* in Sierra Nevada are located in a geographical but not a climatic rear-edge. And the current niche is a result of land-use changes, which further complicates the definition of potential rear-edges. Similarly, high values of resilience showed by other Mediterranean species located in their rear-edge [*e.g.* Pinus sylvestris, @Herrero2014] seem to be related with stabilizing mechanisms promoting community resilience that can buffer the impacts of extreme events.

Greenness and tree-growth of *Q. pyrenaica* showed different sensitivity to severe droughts. During the last drought events EVI was reduced to 90.3 % (Table 1), whereas BAI was reduced to 70.8 % (Table S2), suggesting a lower sensitivity to drought of greenness than of tree-growth, particularly for drier sites. These findings are in accordance with previously works showing that tree-growth is a more sensitive metric of forest resilience than net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018]. Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based (*i.e.* ecosystem or stand based); the growth reduction seems to be more mediated by sink that by source limitations [@Korner2013]. The response of greenness to drought not only reflects the response of trees but from all vegetation within the pixel [@Dionisioetal2012IPoGEC]. In this sense, recent results indicate that remote sensing models underestimate the impact of drought on primary production [@Stocker2019], as soil moisture is not taken into account by this satellite estimates. This is particularly important for *Q. pyrenaica* forests, as soil moisture is a key growth-related factor for this species [@GeaIzquierdo2014]. On the other hand, tree-ring records allow to complement remote sensing data in a longer time scale, by reflecting climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018]. A combined analysis, such the approach used in our study, may allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes [@Babst2017], since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [@GeaIzquierdo2015; @Babst2018].

## Small-scale environmental variability shapes the sensitivity to climate of trees within the rear-edge

Greenness and tree-growth were more affected by drought events in drier northern populations than in more humid southern oak populations, *e.g.* the northern site showed higher browning intensity than the southern sites during the 2005 drought event. In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) in the northern site, can be interpreted as higher sensitivity to drought in drier sites [@GeaIzquierdo2014]. Similar to many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q. pyrenaica* along their distribution range in the Iberian Peninsula [@GeaIzquierdo2014]. Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @GeaIzquierdo2014; @GonzalezGonzalez2014; @GarciaGonzalez2017].

It has been described an effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive, thus high-elevations can be used as local refuge for species in response to climate change [@Bhuyan2017]. It is well known that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM]. Remarkably, the trees at CA-High, which are located around 1900 m. *a.s.l.* and represent today the upper altitudinal limit (treeline) for the species in this southernmost location, showed higher BAI than those located at lower-elevations (CA-Low and SJ) (Figure 4). This shows the high variability in the response to climate exhibited in a small gradient and is especially interesting for the southern sites, which are very close to each other, and overall both constitute the rear-edge for the species. The higher values of tree growth in taller trees that we observed in CA-High site strongly suggest a lower water stress at high-elevation sites [as shown by soil analyses along an elevational gradient in this location @CoboDiaz2017]. Furthermore, as mentioned, the recent growth increase only at the most humid locations a likely concomitant positive effect of a warming climate [@GeaIzquierdo2014]. Finally and in a context of global change, in addition to the withstanding lower water stress, it is likely that the less accessible sites at higher elevations would suffer a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4).

@@

# Conclusions 
Severe drought events provoke important reductions in primary and secondary growth of *Q. pyrenaica* forests in the studied relict, rear-edge forest. However, we observed no negative growth trends despite we had expected vulnerability to climate change for these relict stands. Furthermore, we observed positive trends for primary growth (*i.e.* greenness) at the ecosystem scale and a positive steep trend of secondary growth at the tree level at the most humid site along the climatic gradient analyzed. The trees exhibited high resilience values in response to drought, particularly in the long-term scales, and resilience was directly related to drought severity. These results would be consistent with other research showing that this mountain region is still acting as a refuge for deciduous species, including *Quercus*. The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent and can vary in very small gradients (*i.e.* following ecological thresholds). This is particularly relevant for rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia. The analysis of tree-growth dynamics revealed suppression and release events that were consistent with legacies left by land-use in local forest dynamics, as inferred from an exhaustive review of historical documents. In this sense our results highlight the importance of land-use legacies for highly transformed Mediterranean systems. This is relevant for tree species with a high sensitivity to climate change, such us *Q. pyrenaica*, not only for conservation per se of the species, but for all ecosystem services that these singular forests offer at Mediterranean mountain. Following our study it is necessary to further disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here.

@@

We would like to thank Sammy Lucy Behle, M. Suarez-Muñoz and F.J. Bonet-García for their assistence in the field monitoring. A. Reyes and J. Blanco helped in the filtering of satellite data. F.J. Cano-Manuel and F.J. Navarro provided worthwhile information about oak management projects in Sierra Nevada. We also are very thankful to F.J. Bonet-García for his valuable comments in an earlier version of the manuscript. Environment Department of Andalusian Regional Government and Director of Sierra Nevada National and Natural Park provided permission for field work. AJPL wants to thank the invaluable support received from his family over the years. This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” Project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project). We also thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* and H2020 project *ECOPOTENTIAL: Improving future ecosystem benefits through earth observations* (http://www.ecopotential-project.eu/) (No 641762), for their funding support.

>>>>

# Discussion
## Land-use legacies shape sensitivity to climate change of forests and the present rear-edge

The Mediterranean landscapes have suffered a strong transformation driven by human activities throughout history even at high elevations [@GarciaAlix2017]. The review of historical documents shows how forest clearings, firewood extraction, charcoal exploitations and mining have strongly impacted the forests at Sierra Nevada (Table S4) where in addition it has been estimated a historical loss of about 90% of broadleaf *Quercus* species cover from medium and low elevations [@JimenezOlivencia2015]. Therefore, it seems logic to think that the geographical limit of the current rear-edge and relict stands of the deciduous oak studied does not necessarily match the potential dsitribution limit related to edaphoclimatic constraints (*i.e.* the ecological rear-edge). Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland) [@JimenezOlivencia2015]. It has been described a sequential distribution of historical land uses along the elevational gradient in the northern slopes of Sierra Nevada (*e.g.* SJ site): grasslands and shrublands for cattle farming were located at the highest elevations; then forest stands with some croplands; and at the lowest elevations, irrigated terraces with tree crops [@JimenezOlivencia2015]. In addition, other activities like mining should have modified the forest structure at that northern location where there is with a high concentration of mines and quarries that were exploited intermittently throughout history. On the other hand, woodlands in the southern slopes (*e.g.* CA site) were mixed with a greater percentage of croplands along the elevational gradient where oaks occur [@JimenezOlivencia2015]. Firewood, charcoal extraction and acorns have been resources intensively exploited in southern sites, until at least the middle of the 20th century when these activities sharply decreased mainly due to rural abandonment and use of gas and fossil fuels [@Valbuena2013; @JimenezOlivencia2015].

Some of the previous historical records of land-use were likely expressed by the growth chronologies. The release event in 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral [@Titos1990]. This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth of the remnant trees included in the northern site (Figure S4). The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 5, S4). There are no records of forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest] and no record for any logging  during 1995 - 2000 period (F.J. Cano-Manuel *personal communications*). Therefore this release might be related with a natural drought-induced mortality event after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004]. On the other hand, the strong positive correlations of SPEI with tree-growth for this site show a high sensitivity to water availability [@GeaIzquierdo2014]. For southern sites, the only positive release event showed at the earliest years for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].  In this site we observed an increase in growth especially in high elevation trees, where moisture is less limiting, a pattern which concurs with a forest expansion into marginal abandonded croplads and an increase in tree-cover [@CamachoOlmedo2002].

@@

Severe drought negatively affects both primary and secondary growth of *Q. pyrenaica* forests. This was expressed by the observed reduction in greenness and tree-growth in response to the 2005 and 2012 drought events, and the consistent radial growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014]. Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic precipitation negative anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018]. Although the south of Iberian Peninsula suffered severe drought events in the last decades (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* for the last 16 years, which is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather humid decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto]. For tree-growth, we also observed positive trends in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along its distribution range only at high elevation sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold limited high elevation sites [@Salzer2009; @GeaIzquierdo2014]. Importantly, for rear-edges where climate change is a threat we would have expected negative growth trends as shown for temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017]. Therefore there is much small-scale variability in the response to climate within the rear-edge we had "a priori" considered in our study, where we even observed positive effects of climate change in certain stands. This suggests that the rear-edge needs to be redefined in consequence [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear-edge) and the potential ecological (limiting) rear-edge of species.

Despite that 2012 drought event was more severe and intense than that in 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012. This could be explained by the different timing of the two droughts. The 2012 drought was a winter drought [@Trigo2013] occurring earlier than the 2005 drought and although the latter lasted less than the 2012 drought, it matched the period of maximum growth for *Q. pyrenaica* forests in late spring (Figure S7). Thus, as bserved for other species [*e.g.* Holm oak; @Camarero2015b] this highlights the importance of the timing of drought as a key factor determining the tree recovery after drought [@Huang2018]. For tree-growth, the highest values of resilience were obtained for the two most severe events, namely 1995 and 1999 (Table S3) and we observed a positive relationship between the tree-growth resilience and drought severity (Figures 6). This result seems to contradict our initial hypothesis in which we predicted low resilience values for this oak species in their rear-edge after extreme drought events. Yet, the high resilience values obtained for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality (F.J. Navarro *personal communications*), which would show the influence of stand dynamics partly shadowing interpretation of tree resilience estimated from individual growth series.

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of species [@Hampe2005]. A high vulnerability to drought is often assumed for populations located at their rear-edge [@MartinezVilalta2018]. However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q. pyrenaica* and high values of resilience to severe droughts. Our findings are in agreement with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017]. @MartinezVilalta2018 pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability showed by rear-edge populations. Our results would highlight the ample small-scale variability at the ecological boundary and need to better confine the rear-edge limit in our studied forest. The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013] seem to be indicative of high local adaptation of this oak. In addition the local environmental conditions of the sites where this species inhabits at Sierra Nevada would explain the low vulnerability to drought showed by this oak. It needs to be mentioned that we only studied adult individuals established decades or centuries ago, meaning that it needs to be assessed if the species expresses resilience or to the contrary vulnerability if we analyzed seedling regeneration, likewise many other Mediterranean species at their xeric limit (2-3 CITAS). A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018; @VilaCabrera2019]. In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q. pyrenaica* in Sierra Nevada are located in a geographical but not a climatic rear-edge. And the current niche is a result of land-use changes, which further complicates the definition of potential rear-edges. Similarly, high values of resilience showed by other Mediterranean species located in their rear-edge [*e.g.* Pinus sylvestris, @Herrero2014] seem to be related with stabilizing mechanisms promoting community resilience that can buffer the impacts of extreme events.

## Small-scale environmental variability shapes sensitivity to climate of trees within the rear-edge

Since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [@GeaIzquierdo2015; @Babst2018] a combined analysis as in our study allows to better disentangle the small and large-scale variability in growth induced by climate from that driven by community-level ecological processes [@Babst2017]. Greenness exhibited a lower sensitivity to drought than tree-growth, particularly for drier sites. These findings are in accordance with previously works showing that tree-growth is a more sensitive metric of forest resilience than net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018] and would suggest that the growth reduction could more mediated by sink that by source limitations [@Korner2013; @Fatichi2014]. On the other hand, tree-ring records allow to complement remote sensing data in a longer time scale, by reflecting climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].

Greenness and tree-growth were more affected by drought events in drier northern populations than in more humid southern oak populations, *e.g.* the northern site showed higher browning intensity than the southern sites during the 2005 drought event and the stronger correlations of tree-growth with SPEI (Hydrological and summer) in the northern site can be interpreted as higher sensitivity to drought in drier sites [@GeaIzquierdo2014]. Similar to many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q. pyrenaica* along their distribution range in the Iberian Peninsula [@GeaIzquierdo2014]. Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @GeaIzquierdo2011; @GarciaGonzalez2017].

Higher elevations within warm and dry ecosystems can help species alleviate climatic stress and 
increase resistance to drought and stands of several tree-species located at higher elevations were less drought sensitive. Thus high-elevations are generally considered as potential local refuge for species in response to climate change [@Bhuyan2017]. It is well known that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM]. Trees at CA-High sowhed higher BAI than those located at lower-elevations (CA-Low and SJ) (Figure 4). This shows the high variability in the response to climate exhibited in a small gradient, which is especially interesting for southern sites, which are very close to each other, and overall both constitute the rear-edge for the species. The higher values of tree growth in taller trees that we observed in CA-High site strongly suggests lower water stress at high-elevation sites [as shown by soil analyses along an elevational gradient in this location @CoboDiaz2017]. Furthermore, as mentioned, the recent growth increase only at the most humid locations a likely concomitant positive effect of a warming climate [@GeaIzquierdo2014]. Finally and in a context of global change, in addition to withstanding lower water stress, it is likely that the less accessible sites at higher elevations would suffer a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4). Overall, our results show how management history conditions tree-growth and resilience to climate change of tree species, highlighting the importance of land-use legacies in Mediterranean forests [@Navarro2013; @DoblasMiranda2017] which, hence, will also strongly determine the current geographical distribution of marginal stands, regardless of the potential extent of ecological marginality of species.

@@

# Conclusions 
The ecological and geographical read-edges did not seem to fully match in our study. Severe drought events provoke important reductions in primary and secondary growth of *Q. pyrenaica* forests in the studied relict, rear-edge forest. However, we observed no negative growth trends despite we had expected vulnerability to climate change for these relict stands. Furthermore, we observed positive trends for primary growth (*i.e.* greenness) at the ecosystem scale and a positive steep trend of secondary growth at the tree level at the most humid site along the climatic gradient analyzed. The trees exhibited high resilience values in response to drought, particularly in the long-term scale, and resilience was directly related to drought severity. These results would be consistent with other research showing that this mountain region is still acting as a refuge for deciduous species, including *Quercus*. The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent and can vary in very small gradients (*i.e.* following ecological thresholds). This is particularly relevant for rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia. The analysis of tree-growth dynamics revealed suppression and release events that were consistent with legacies left by land-use in local forest dynamics, as inferred from an exhaustive review of historical documents. In this sense our results highlight the importance of land-use legacies for highly transformed Mediterranean systems. This is relevant for tree species with a high sensitivity to climate change, such us *Q. pyrenaica*, not only for conservation per se of the species, but for all ecosystem services that these singular forests offer at Mediterranean mountain. Following our study it is necessary to further disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here. Since we only studied resilience of already established adult trees, monitoring should be enlarged to analyze overall species resilience including that of seedling regeneration. Our results show how management history conditions tree-growth and resilience to climate of tree species. They highlight the importance of land-use legacies and how they affect the current geographical distribution of marginal stands, which therefore can be partly disconnected from actual ecologically marginal conditions.

@@

We would like to thank Sammy Lucy Behle, M. Suárez-Muñoz and F.J. Bonet-García for their assistence in the field monitoring. A. Reyes and J. Blanco helped in the filtering of satellite data. F.J. Cano-Manuel and F.J. Navarro provided worthwhile information about oak management projects in Sierra Nevada. We also are very thankful to F.J. Bonet-García for his valuable comments in an earlier version of the manuscript. AJPL wants to thank the invaluable support received from his family over the years. Environment Department of Andalusian Regional Government and Director of Sierra Nevada National and Natural Park provided permission for field work. This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” Project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project). We also thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* and H2020 project *ECOPOTENTIAL: Improving future ecosystem benefits through earth observations* (http://www.ecopotential-project.eu/) (No 641762), for their funding support.
